R and Python clients for GBIF species occurrence data

### 2016-03-11 manuscript.Rmd
converted ms to discuss python in addition to R client

---
layout: review, 11pt
linenumbers: true
title: "rgbif: R client for working with GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

@@

# The rgbif package

@@

## Package interface

`rgbif` is designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, taxonomic names,
occurrences, and maps. We also include functions to interface with the OAI-PMH
GBIF service; only dataset information is available vis this service, however. We
ignore maps in `rgbif` as it is concerned with generating maps primarily for web
applications. `rgbif` has a suite of functions dealing with each of registry, taxonomic
names, and occurrences - we'll go through each in turn describing design and example usage.

## GBIF feedback loop

With each request `rgbif` makes to GBIF's API, we send request headers that
tell GBIF that the request is coming from `rgbif`, including what version of
the package. This helps GBIF know what proportion of requests are coming from
this package, and therefore R, as most requests likely will come from `rgbif`;
this information is helpful for them in thinking about how people are using
GBIF data.

@@

The GBIF registry API services are spread across five sets of functions
via the main API:

@@

And dataset information in general is available via the OAI-PMH service,
functions in `rgbif` prefixed with `gbif_oai_`.

@@

See also `datasets()` and `dataset_suggest()` for searching for datasets.

@@

@@

Networks, nodes and installations are at a higher level of organization
above datasets, but can be useful if you want to explore data from given
organizations. Here, we search for the first 10 GBIF networks, returning
just the title field.

```{r}
networks(limit=10)$data$title
```

@@

The GBIF taxonomic names API services are spread across five functions:

@@

@@

We can search the backbone taxonomy with the function `name_backbone()`. Here,
we're searching for the name _Poa_, restricting to genera, and the family
_Poaceae_.

@@

@@

@@

GBIF provides two ways to get occurrence data: through the
`/occurrence/search` route (see `occ_search`), or via the
`/occurrence/download` route (many functions, see below).
`occ_search()` is the main function for the search route,
and is more appropriate for smaller data, while
`occ_download*()` functions are more appropriate for larger
data requests.

@@

We think the search service is still quite useful for many people
even given the 200,000 limit. For those that need more data, we have
created a similar interface in the `download_*()` functions, that should
be easy to use. Users should take note that using the download service
has a few extra steps to get data into R, but is straight-forward.

@@

```{r}
library("dplyr")
dat <- occ_download_import(res)
dat %>%
  select(gbifID, decimalLatitude, decimalLongitude)
```

@@

The downloaded dataset from GBIF is actually a Darwin Core Archive (DwC-A), an
internationally recognized biodiversity informatics standard
(http://rs.tdwg.org/dwc/). The DwC-A downloaded is a compressed folder with
a number of files, including metadata, citations for each of the datasets included
in the download, and the data itself, in separate files for each dataset as well
as one single `.txt` file. In `occ_download_import()`, we simply fetch data from the
`.txt` file. If you want to dig into the metadata, citations, etc., it is easily
accessible from the folder on your computer.

@@

* Get a single numeric count of occurrences - `occ_count()`
* Search for occurrences - `occ_search()`
* A simplified and optimized version of `occ_search()` - `occ_data()`
* Get occurrences by occurrence identifier - `occ_get()`
* Get occurrence metadata - `occ_metadata()`

@@

@@

@@

@@

The `rgbif` R package provides a programmatic interface to GBIF's
application programming interface (API) - a powerful tool for making
research using species occurrence data reproducible. In fact,
the `rgbif` package has already been used in 22 scholarly publications
(as of 2015-11-14).

@@

@@

[mit]: http://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: http://www.gbif.org/developer/summary
[gbifissues]: http://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html
[taxize]: https://github.com/ropensci/taxize
[spocc]: https://github.com/ropensci/spocc
[mapr]: https://github.com/ropensci/mapr

>>>>

---
layout: review, 11pt
linenumbers: true
title: "R and Python clients for GBIF species occurrence data"
author:
  - name: Scott Chamberlain
    affiliation: cstar
    email: scott(at)ropensci.org
    footnote: Corresponding author
  - name: Carl Boettiger
    affiliation: boettig
    email: carl(at)ropensci.org
address:
  - code: cstar
    address: |
      University of California, Berkeley, CA, USA
  - code: boettig
    address: |
      University of California, Berkeley, CA, USA
abstract: |
      1. xxx

@@

# The clients

## The rgbif package

@@

## The pygbif library

`pygbif` [@pygbif] is a Python library for working with GBIF data in
the Python programming environment [@python]. Python is a general
purpose programming language used widely in all sectors, and for
all parts of software development including server and client
side use cases. Python is used exclusively in some scientific
disciplines (e.g., astronomy), and has partial usage in other
disciplines. A Python client for GBIF data is an important tool
given the even wider usage of Python than R, though maybe slightly
less than R for ecology/biology.

```{python eval=FALSE}
pip install pygbif
```

```{python eval=FALSE}
import pygbif
```

The `pygbif` library is less mature and complete than the R package. It also
uses an [MIT license][mit] to maximize use everywhere. `pygbif` is developed
publicly on GitHub at [https://github.com/sckott/pygbif][ghpygbif], where
development versions of the package can be installed, and bugs and feature
requests reported. Stable versions of `pygbif` can be installed from
[pypi][pypipygbif], the distribution network for Python libraries.

### Library interfaces

`rgbif` and `pygbif` are designed following the [GBIF Application Programming Interface][gbifdocs],
or API. The GBIF API has four major components: registry, taxonomic names,
occurrences, and maps. We also include functions to interface with the OAI-PMH
GBIF service; only dataset information is available vis this service, however. We
ignore maps in both libraries as it is concerned with generating maps primarily for web
applications. Both libraries have a suite of functions dealing with each of registry, taxonomic
names, and occurrences - we'll go through each in turn describing design and example usage.

## GBIF headers

With each request `rgbif` and `pygbif` make to GBIF's API, we send request headers that
tell GBIF what client the request is coming from, including what version of
the library. This helps GBIF know what proportion of requests are coming from
which client, and therefore from R vs. Python; this information is helpful for GBIF
in thinking about how people are using GBIF data.

@@

The GBIF registry API services are spread across five sets of functions
via the main GBIF API:

@@

Dataset information in general is available via the OAI-PMH service,
functions in `rgbif` prefixed with `gbif_oai_`, but not available in `pygbif`
yet.

@@

See also `datasets()` and `dataset_suggest()` in `rgbif` for searching for datasets.

In Python, we can similary search for datasets. Here, search for datasets of type
`OCCURRENCE`:

```{python eval=FALSE}
from pygbif import registry
registry.datasets(type="OCCURRENCE")
```

@@

And in Python, get metrics for the same dataset as above:

```{python}
from pygbif import registry
registry.dataset_metrics(uuid='ec93a739-1681-4b04-b62f-3a687127a17f')
```

@@

Networks, nodes and installations are at a higher level of organization
above datasets, but can be useful if you want to explore data from given
organizations. Here, in R we search for the first 10 GBIF networks, returning
just the title field.

```{r}
networks(limit = 10)$data$title
```

And in Python:

```{python}
from pygbif import registry
registry.networks(limit = 10)
```

@@

The GBIF taxonomic names API services are spread across five functions in `rgbif`:

@@

`pygbif` only has `name_backbone()` and `name_suggest()` at this time.

@@

We can search the backbone taxonomy with the function `name_backbone()` in both
R and Python clients. Here, we're searching for the name _Poa_, restricting to
genera, and the family _Poaceae_, in R

@@

and in Python

```{python eval=FALSE}
from pygbif import species
res = species.name_backbone(name='Poa', rank='genus', family='Poaceae')
[ res[x] for x in ['usageKey', 'kingdom'] ]
```

@@

The same in Python

```{python eval=FALSE}
from pygbif import species
species.name_suggest(q='Pum', limit = 6)
```

@@

GBIF provides two ways to get occurrence data: through the
`/occurrence/search` route (see `occ_search` in `rgbif`, or
`occurrences.search` in `pygbif`), or via the
`/occurrence/download` route (many functions, see below).
`occ_search()`/`occurrences.search` is the main function for the search route,
and is more appropriate for smaller data, while
the download functions are more appropriate for larger
data requests.

@@

We think the search service is still quite useful for many people
even given the 200,000 limit. For those that need more data, we have
created a similar interface in the download functions, that should
be easy to use. Users should take note that using the download service
has a few extra steps to get data into R, but is straight-forward.

@@

```{r eval=FALSE}
library("dplyr")
dat <- occ_download_import(res)
dat %>%
  select(gbifID, decimalLatitude, decimalLongitude)
#>       gbifID abstract accessRights accrualMethod accrualPeriodicity accrualPolicy alternative audience
#> 1  725767384       NA                         NA                 NA            NA          NA       NA
#> 2  725767447       NA                         NA                 NA            NA          NA       NA
#> 3  725767450       NA                         NA                 NA            NA          NA       NA
#> 4  725767513       NA                         NA                 NA            NA          NA       NA
#> 5  725767546       NA                         NA                 NA            NA          NA       NA
#> 6  725767579       NA                         NA                 NA            NA          NA       NA
#> 7  725767609       NA                         NA                 NA            NA          NA       NA
#> 8  725767645       NA                         NA                 NA            NA          NA       NA
#> 9  725767678       NA                         NA                 NA            NA          NA       NA
#> 10 725767681       NA                         NA                 NA            NA          NA       NA
#> ..       ...      ...          ...           ...                ...           ...         ...      ...
#> Variables not shown: available (lgl), bibliographicCitation (chr), conformsTo (lgl), contributor (lgl),
#>      coverage (lgl), created (chr), creator (lgl), date (lgl), dateAccepted (lgl), dateCopyrighted
#>      (lgl), dateSubmitted (lgl), description (lgl), educationLevel (lgl), extent (lgl), format (lgl),
#>      hasFormat (lgl), hasPart (lgl), hasVersion (lgl), identifier (chr), instructionalMethod (lgl),
```

In Python

```{python eval = FALSE}
from pygbif import occurrences as occ
occ.download('taxonKey = 3119195')
(res = occ.download_get("0000840-150615163101818", overwrite = True))
```

We don't have `pygbif` functionality at the moment for importing data, but
it's coming soon.

@@

The downloaded dataset from GBIF is a Darwin Core Archive (DwC-A), an
internationally recognized biodiversity informatics standard
([http://rs.tdwg.org/dwc/](http://rs.tdwg.org/dwc/)). The DwC-A downloaded is a
compressed folder with a number of files, including metadata, citations for each
of the datasets included in the download, and the data itself, in separate files for
each dataset as well as one single `.txt` file. In `rgbif::occ_download_import()`, we
simply fetch data from the `.txt` file. If you want to dig into the metadata,
citations, etc., it is easily accessible from the folder on your computer.

@@

* Get a single numeric count of occurrences - `rgbif`: `occ_count()` / `pygbif`: `occurrences.count`
* Search for occurrences - `rgbif`: `occ_search()` / `pygbif`: `occurrences.count`
* A simplified and optimized version of `rgbif`: `occ_search()` or `occ_data()` / `pygbif`: `occurrences.count`
* Get occurrences by occurrence identifier - `rgbif`: `occ_get()` / `pygbif`: `occurrences.count`
* Get occurrence metadata - `rgbif`: `occ_metadata()` / `pygbif`: `occurrences.count`

@@

__R__

@@

__Python__

The equivalent occurrence search workhorse in `pygibf` is `occurrences.search`.

```{python eval=FALSE}
from pygbif import species
from pygbif import occurrences as occ
key = species.name_suggest(q = 'Helianthus annuus', rank = 'species')['data'][0]['key']
occ.search(taxonKey = key, limit = 2)
```

```{python eval=FALSE}
occ.search(scientificName = 'Ursus americanus')
```

```{python eval=FALSE}
from pygbif import species
from pygbif import occurrences as occ
splist = ['Cyanocitta stelleri', 'Junco hyemalis', 'Aix sponsa']
keys = [ species.name_suggest(x)['data'][0]['key'] for x in splist ]
occ.search(taxonKey = keys, limit = 5)
```

```{python eval=FALSE}
from pygbif import occurrences as occ
# well known text
occ.search(geometry = 'POLYGON((30.1 10.1, 10 20, 20 40, 40 40, 30.1 10.1))', limit = 20)
# bounding box
occ.search(geometry = '-125.0,38.4,-121.8,40.9', limit = 20)
```

```{python eval=FALSE}
from pygbif import occurrences as occ
occ.search(hasCoordinate = True, limit = 5)
```

```{python eval=FALSE}
from pygbif import occurrences as occ
occ.search(hasGeospatialIssue = True, limit = 5)
```

@@

There's no equivalent interface in `pygbif`.

@@

The `rgbif` and `pygbif` libraries provide programmatic interfaces to GBIF's
application programming interface (API) - a powerful tool for making
research using species occurrence data reproducible. In fact,
the `rgbif` package has already been used in 22 scholarly publications
(as of 2015-11-14).

@@

The `pygbif` library is still in early development, and will
greatly benefit from any feedback and use cases.

@@

[mit]: http://choosealicense.com/licenses/mit/
[ghrgbif]: https://github.com/ropensci/rgbif
[cranrgbif]: https://cran.rstudio.com/web/packages/rgbif/
[gbifdocs]: http://www.gbif.org/developer/summary
[gbifissues]: http://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html
[taxize]: https://github.com/ropensci/taxize
[spocc]: https://github.com/ropensci/spocc
[mapr]: https://github.com/ropensci/mapr
[ghpygbif]: https://github.com/sckott/pygbif
[pypipygbif]: https://pypi.python.org/pypi/pygbif
